---
title: "Finding Numeric Regressions with daffiz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finding Numeric Regressions with daffiz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`daffiz` compares numeric measures in two tables that should be nearly
identical. Every diagnostic derives from one long cell table, and every
difference retains the original row number from each input.

This vignette follows the usual investigation from the overall result to the
responsible measures, source records, and exact numeric cells.

```{r setup}
library(daffiz)

baseline <- data.frame(
  id = 1:4,
  region = c("north", "south", "east", "west"),
  amount = c(100, 200, 300, 400),
  score = c(0.5, 0.6, 0.7, 0.8)
)

candidate <- data.frame(
  id = 2:5,
  region = c("south", "east", "west", "central"),
  amount = c(200.02, 330, 400, 500),
  score = c(0.6, 0.72, 0.8, 0.9)
)

cmp <- compare_dt(
  baseline,
  candidate,
  abs_tol = c(.default = 0, amount = 0.05)
)
```

## 1. Do the measures match?

```{r matching}
is_matching(cmp)
summary(cmp, max_cells = 3)
```

The report names the inferred identity and measure columns, the tolerance
policy, unmatched-row counts, and bounded samples. Here, `id` and `region` form
the identity because they are not ordinary double columns. `amount` and `score`
are measures.

If those defaults are inappropriate, select the roles explicitly:

```{r roles}
compare_dt(
  baseline,
  candidate,
  by = "id",
  compare = c("amount", "score"),
  abs_tol = c(.default = 0, amount = 0.05)
)
```

## 2. Which measures differ?

```{r columns}
column_summary(cmp)
diff_columns(cmp)
```

`column_summary()` includes exact cell counts, mismatch fractions, signed and
absolute mean differences, RMSE, maximum absolute difference, and the 95th
percentile absolute difference. `diff_columns()` keeps only affected measures
and ranks the largest problems first.

## 3. Which source records differ?

```{r rows}
diff_rows(cmp)
diff_rows(cmp, view = "paired")

diff_indices(cmp, "x")
original_rows(cmp, "x")

x_only(cmp)
y_only(cmp)
```

`.row_x` and `.row_y` always refer to positions in `baseline` and `candidate`
before any sorting or melting. `original_rows()` uses private snapshots, so the
recovered records retain the original column order, classes, and values even if
a caller later modifies a `data.table` by reference.

## 4. What is the exact shape of the differences?

```{r cells}
diff_cells(cmp)
diff_cells(cmp, columns = "amount")
```

Each row is one numeric cell classified as `diff`, `x_only`, or `y_only`.
`all_cells()` also includes equal cells, which supply the exact denominators for
the summaries.

Duplicated identities need an explicit policy. The default `"pair"` policy
sorts and pairs duplicated records deterministically but warns that the pairing
is heuristic. `"error"` stops. `"report"` records and excludes ambiguous
groups; because those records were not compared, the resulting comparison is
incomplete and `is_matching()` returns `FALSE`.

## 5. Where are differences concentrated?

The optional plotting layer is a single view over the same cell table: an
Amelia-style difference map, in the spirit of `Amelia::missmap()`. Observations
run down the y axis, numeric measures across the x axis, green where the two
tables agree and red where they do not.

```{r difference-map, fig.width = 7, fig.height = 4}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_diff(cmp)
}
```

Rows keep their original source order rather than being ranked by how affected
they are, so a run of regressions in one region of the input reads as a band.
Above `max_rows` records the map aggregates contiguous rows into bins and
shades each tile by the share of its cells that differ; no row is dropped, and
the subtitle discloses the binning.

The fill colours are the Okabe-Ito bluish green and vermillion, which read as
green and red but stay distinguishable under the common forms of colour
blindness, and each tile carries a symbol while the map is small enough to
label it, so the classification never rests on colour alone.

`plot_data()` returns the exact table behind the map. It works without
`ggplot2`, which is useful for audits and custom graphics.

```{r plot-data}
plot_data(cmp)
```

Measure- and record-level questions are answered by the data accessors above
rather than by further chart types.

## Regression tests

`expect_dt_equal()` exposes the same semantics inside testthat:

```{r expectation}
expect_dt_equal(baseline, baseline)
```

Failure messages contain bounded column, row, and cell diagnostics and carry
the complete comparison object as a condition attribute.

## Time and memory

The canonical result has approximately `n_rows * n_measures` rows. Both inputs
are melted and joined in memory, and equal cells are retained for exact summary
denominators. Narrow very wide comparisons with `compare` or `exclude`.

For wide tables, `batch` limits how many measures are melted and joined at one
time while preserving the complete result. For example,
`compare_dt(x, y, batch = 10)` processes ten measures per batch. Batching lowers
the peak size of construction intermediates; it does not reduce the final
canonical cell table.

Benchmarks measured about 64 retained bytes per unique-key cell and much higher
working memory during construction. A projected result above 10 million cells
therefore warns by default. Set a different warning threshold with
`options(daffiz.max_cells = ...)` after considering the input width, batch size,
and available memory.
